PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-10 IEEEtran/paper.tex
Update for major revision

\definecolor{lightred}{rgb}{1.0,0.6,0.6}
\definecolor{lightgreen}{rgb}{0.6,1.0,0.6}

\DeclareRobustCommand{\add}[1]{{\sethlcolor{lightgreen}\hl{#1}}}
\DeclareRobustCommand{\remove}[1]{{\sethlcolor{lightred}\hl{#1}}}

\DeclareRobustCommand{\addCite}[1]{{\sethlcolor{lightgreen}\hl{\mbox{#1}}}}
\DeclareRobustCommand{\removeCite}[1]{{\sethlcolor{lightred}\hl{\mbox{#1}}}}

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\addCite{\cite{higham2002accuracy}}, which require detailed modeling of the program and its
numerical implementation. Similarly, static code analysis techniques~\addCite{\cite{goubault2006static}} hardly
scale to large code bases, particularly in dynamically-typed languages such as
Python.

@@

\pytracer detects numerical instabilities by computing summary statistics across
multiple executions perturbed with numerical noise. While \pytracer's
instrumentation, trace aggregation, and visualization work with various types of
numerical noise, we experimented primarily with Monte-Carlo Arithmetic (MCA) as
it is an accurate model for floating-point errors~\addCite{\cite{verificarlo}}.

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis [23], and random
seed analysis [11].  \hl{PyTracer is agnostic to the noise model used and therefore
    works for all methods mentioned}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision. MCA
can be applied in three modes: Random Rounding (RR), Precision Bounding (PB),
and full MCA, which apply function $inexact$ to the output, the inputs, or both,
respectively. While the RR mode is equivalent to stochastic rounding, the PB
mode can also identify catastrophic cancellations.

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA.  
Sohier et al.~\cite{sohier2018confidence} recently provided a generalization of
this formula to include confidence intervals.

@@

\pytracer stores traces in the pickle format, a binary format to serialize
Python objects. The main advantages of the pickle format compared to other
serialization approaches such as marshal or JSON are its ability to serialize
most Python objects and its native compression.  In addition, pickle writes
Python objects sequentially, which preserves the temporal ordering required by
subsequent trace analyses.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state. 
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from a different control flow path
    and will stop the parsing when encountering one.}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figure/pytracer_layout.pdf}
    \caption{Pytracer visualization layout.}
    \label{fig:visu-layout}
\end{figure}

@@

\section{Numerical evaluations}

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy, SciPy, and
scikit-learn. \add{Only instrumented the Python interpreter would lead to
    only perturbe pure Python floating-point operations and thus to miss
    operations in compiled codes (i.e. Cython and shared libraries).}
\hl{We repeated the experiment with two MCA modes, Random Rounding
    (RR) and Full MCA, resulting in 2 experiments for each example}. We used the
default virtual precision in Verificarlo, namely 24 bits for single-precision
floating-point values and 53 bits for double-precision values, which simulates
machine error. We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and
scikit-learn-0.23.2.

\subsection{Results overview}

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with three possible outcomes: 1) green for
success, meaning that the execution ended well and no issues were raised by the
application, 2) red for numerical error in the application, meaning that the
execution raised a Python error before the end, and 3) orange for errors due to
the fuzzy noise injection that are not considered numerical errors, meaning that
the execution raised a Python error but due to a perturbation that should not
have been introduced (i.e. a floating-point value representing an integer).
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples and showed an average
precision loss of 8 bits. The remaining examples failed due to numerical errors
in the libraries, which we discuss hereafter. The instrumentation with Full MCA
is more invasive and reduced the number of successful executions to 12/40. Among
the 28/40 failed executions, one failed due to an actual numerical error
(\texttt{Bayesian Ridge Regression} example) and the other ones failed due to
perturbations that should not have been introduced.

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the precision for the main outputs of SciPy and scikit-learn examples. We
measured the precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produces many outputs, the average value among all of them. The numerical
quality across all examples varies from 0 to 52, with an average of 44
significant bits. For both modes, we note that only two examples failed due to a
numerical error (\texttt{face\_recognition} for RR and \texttt{Bayesian Ridge
    Regression} for Full MCA).

@@

        % \tristan{Replace FFT with SciPy FFT, Interpolation with SciPy Interpolation and Optimization with SciPy Optimization, or add a SciPy header to these rows.} \tristan{Nitpick: you could add a color scheme to sigbits (for instance shades of blue), to make the column more easily readable.} \\
        \emph{trace}: outcome of \pytracer tracing, \emph{parse}: outcome of \pytracer trace parsing and aggregation, \emph{result}: precision for main outputs in significant bits. \valid: successful run, \warn: error from invalid noise injection, \cross: numerical error or divergent traces. \textit{nan} stands for (\texttt{Not-A-Number}). Example names are linked to their source code.}

@@

\subsection{Effect of MCA modes}
\label{sec:impact_mca_modes}
Random Rounding perturbs only the result of a floating-point operation while
Full MCA perturbs both the inputs and the result. Therefore, Full MCA may
trigger catastrophic cancellations while Random Rounding only introduces
round-off errors which generally have a lesser impact. Moreover, Random Rounding
preserves exact operations --- operations where the result can be exactly
represented at the virtual precision --- while Full MCA does not.

\pytracer can only aggregate traces obtained from the same control flow path. In
particular, program branches triggered by an unstable floating-point comparison
lead to different control flows and result in parsing errors. For example,
comparing a residual error or the distance between two points to a strict
threshold in an iterative scheme generally results in unstable branches. Full
MCA even perturbs exact values such as integers represented with floating-point
values, leading to many execution failures. For instance, the \texttt{fft1}
example from SciPy raised an \texttt{ValueError: operands could not be broadcast
    together with shapes (601,) (600,)} error due to an array shape mismatch. A
closer inspection showed that one of the arrays involved in the error was
created with the NumPy function \texttt{linspace} that itself calls function
\texttt{arange} (Listing~\ref{fig:pyarray_range_code}). Lines 3174-3178 show
that the function stores the array size in a floating-point value that will be
perturbed by Full MCA, as shown in Table~\ref{tab:mca_result_linspace},
resulting in array sizes that fluctuate between N and N+1. It is worth noting
that 78\% of execution issues encountered in Full MCA mode instrumentation are
due to this type of ``typing" errors.  
Those code regions should not be perturbed by MCA since the injected noise does
not make sense.

\begin{listing}
    \begin{lstlisting}[language=C,style=customC,numbers=left, firstnumber=3163, mathescape=true]
/* NUMPY_API Arange, */
NPY_NO_EXPORT PyObject *
PyArray_Arange(double start, double stop, double step, int type_num) {
    npy_intp length; 
    PyArrayObject *range; $\lstsetnumber{\ldots}$
    ...$\lstresetnumber\setcounter{lstnumber}{3173}$
    double delta, tmp_len; $\lstsetnumber{\ldots}$
    ...$\lstresetnumber\setcounter{lstnumber}{3176}$
    delta = stop - start; 
    tmp_len = delta/step; $\lstsetnumber{\ldots}$
    ...$\lstresetnumber\setcounter{lstnumber}{3189}$
        length = _arange_safe_ceil_to_intp(tmp_len); $\lstsetnumber{\ldots}$
    ...$\lstresetnumber\setcounter{lstnumber}{3201}$
    range = (PyArrayObject *)PyArray_New(&PyArray_Type, 1, &length, type_num, NULL, NULL, 0, 0, NULL);
    \end{lstlisting}
    \caption{Source code of the \texttt{PyArray\_Range} Cython function called by NumPy function \texttt{linspace} to create an array of equally-spaced elements. The temporary array size assigned in line 3178 is stored as a floating-point value and is therefore perturbed in Full MCA mode, leading to differences amplified by the use of ceil rounding at line 3190 and resulting in different array sizes across MCA-perturbed executions. See also Table~\ref{tab:mca_result_linspace}.}
    \label{fig:pyarray_range_code}

\end{listing}

\begin{table}[t]
    \centering
    % \begin{tabular}{ccccccc}
    %     \thickhline
    %     \textbf{Mode} & \textbf{start} & \textbf{stop} & \textbf{step} & \textbf{delta}                & \textbf{tmp\_len} & \textbf{ceil(tmp\_len)} \\
    %     \hline
    %     Exact         & 0              & 600           & 1             & 600                           & 600               & 600                     \\
    %     RR            & 0              & 600           & 1             & 600 $\pm$ 0.0                 & 600 $\pm$ 0.0     & 600 $\pm$ 0.0           \\
    %     MCA           & 0              & 600           & 1             & 599.9999999999999 $\pm$ 8e-14 & 600.0 $\pm$ 8e-14 & 600.01 $\pm$ 0.0995     \\
    %     \thickhline        
    % \end{tabular}
    \begin{tabular}{l|ccc}
        \hline                                                   
        \textbf{Mode}           & Exact & RR            & MCA                   \\
        \hline
        \texttt{start}          & 0     & 0             & 0                     \\
        \texttt{stop}           & 600   & 600           & 600                   \\
        \texttt{step}           & 1     & 1             & 1                     \\ 
        \texttt{delta}          & 600   & 600 $\pm$ 0.0 & 599.9...9 $\pm$ 8e-14 \\
        \texttt{tmp\_len}       & 600   & 600 $\pm$ 0.0 & 600.0 $\pm$ 8e-14     \\
        \texttt{ceil(tmp\_len)} & 600   & 600 $\pm$ 0.0 & 600.01 $\pm$ 0.0995   \\
        \thickhline        
    \end{tabular}
    \caption{Results obtained through lines 3177-3190 in Listing~\ref{fig:pyarray_range_code}. In contrast with RR, Full MCA does not preserve exact operations which leads to an array length oscillating between 600 and 601.}
    \label{tab:mca_result_linspace}
\end{table}

\subsection{SciPy}
\label{sec:scipy_tests}

We tested \pytracer on SciPy examples provided in the tutorial website section.
SciPy is organized in several libraries targeting specific computing domains. We
focused on \texttt{fftpack} (Fast Fourier Transform routines),
\texttt{interpolate} (interpolation and smoothing splines), and
\texttt{optimize} (Optimization and root-finding routines).

\subsubsection{FFT}

This package has three examples : \texttt{discrete\_cosine\_transform},
\texttt{fft\_1D}, and \texttt{fft\_2D}. All computations are done in double
precision. The results for \texttt{discrete\_cosine\_transform},\texttt{fft\_1D}
and  \texttt{fft\_2D} show precise numerical results with 48 significant bits on
average. We observe that the FFT computation is stable to noisy inputs.
Figures~\ref{fig:fft1D_inputs} (inputs) and~\ref{fig:fft1D_outputs} (outputs)
show the mean and standard deviation of the MCA samples. The three columns
correspond to the FFT computations of: 
\begin{enumerate}
    \item $z_1 = \sin(50 . 2\pi . x_i) + \dfrac{1}{2} \sin(80 . 2\pi . x_i),\; x_i = \frac{i}{600},\; i \in [0,600]$
    \item $\mathrm{blackman}(400) \times z_1$
    \item $ z_2= e^{50 . i 2\pi . x_i} + \dfrac{1}{2} e^{-80 . i2\pi .x_i },\; x_i = \frac{i}{800},\; i \in [0,400] $
\end{enumerate}

\add{The} top row in Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs} show the
input mean value while the bottom row shows the standard deviation. The points
with low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs near 0
when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in \mathbb{Z}$.
Indeed, since MCA introduces a slight noise, the result is not exactly 0. We can
see in Figure~\ref{fig:fft1D_inputs} that the maximal noise introduced by MCA on
the inputs is in the order of $10^{-14}$, two orders of magnitude higher than
the $ulp \simeq 10^-16$ for double precision. This slight noise introduced can
be interpreted as white noise on the input signal. We can see on the bottom row
of Figure~\ref{fig:fft1D_outputs} that the frequency noise is of the same order
of of magnitude as the input noise, which is expected. The two peaks at x=38 and
x=562 corresponds to the actual frequencies of the input signal.

@@

\subsection{Interpolation}

This package has five examples: \texttt{interpolation\_1D},
\texttt{multivariate\_data}, \texttt{bspline}, \texttt{spline\_1D} and
\texttt{spline\_2D}.

\texttt{interpolation\_1D} interpolates function $f(x)=\cos(\frac{-x^2}{9})$
with 11 points $x\in[0,10]$. For the five interpolation methods tested (linear,
cubic, nearest, previous, next), the solution found is precise, up to 51 bits
out of the 53 available.

\texttt{multivariate\_data} interpolates the grid $f(x,y)=x(1-x)\cos(4\pi.x)
    \sin(4\pi.y^2)^2$ with $(x,y) \in [0,1] \times [0,1]$. It samples 1,000 points
for each coordinate and uses three interpolation methods: nearest, linear and
cubic. Our results show a precision of 51 bits on average for the three methods.

\texttt{bspline} compares two edge filters: \texttt{bisplev} evaluates a
bivariate B-spline and its derivative whereas \texttt{convolved2d} convolves two
2-dimensional arrays. Figure~\ref{fig:bspline_rr} shows the significant bits,
mean, and standard deviation maps for the two methods as well as the input image
(Fig.~\ref{fig:bspline_original_image}). The middle row shows \texttt{sepfir2d}
results and the bottom row shows \texttt{convol2d} results. Both methods exhibit
similar precision, with 11 significant bits on average. Standard deviations maps
on Figures~\ref{fig:bspline_bisplev_std} and~\ref{fig:bspline_convol2d_std} show
that regions with low spatial frequencies correspond to regions with low
standard deviation, similar to the FFT results. The SciPy tutorial mentions that
\texttt{bisplev} is faster than \texttt{convol2d}. The comparable numerical
precision observed in our experiments reinforce the use of \texttt{bisplev} over
\texttt{convol2d} in this example.

\texttt{spline\_1D} computes the 1D spline interpolation for $f(x)=\sin(x)$ on
points $x=\frac{2\pi k}{8}, k \in [0, 10]$. The spline interpolation uses
\texttt{splrep} to build the spline representation and \texttt{splrev} to
evaluate the spline at any point. It also computes the integral by using
\texttt{splint} and the root finder \texttt{sproot}. Although the traces
diverge, we can analyze the partial aggregation for the \texttt{splrep},
\texttt{splev} and \texttt{splint}. The three functions show results with an
average precision of 51 bits with RR mode.

\texttt{spline\_2D} computes the 2D spline interpolation for function
$z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in [-1,1]\times[-1,1]$ and a sampling of
21 points for each coordinate. \texttt{bisplrep} builds the spline
representation with the $21 \times 21$ points of $z$ over 71 grid sampling
points for $(x,y)$. Figure~\ref{fig:spline2d_rr} shows the results of the spline
evaluation with \texttt{bisplev}.  We observe a significant loss of precision in
places, which follows an interesting pattern that could be further investigated.

>>>>

\definecolor{lightred}{rgb}{0.6,0.0,0.0}
\definecolor{lightgreen}{rgb}{0.8,1.0,0.8}

\DeclareRobustCommand{\add}[1]{#1}
\DeclareRobustCommand{\remove}[1]{}

\DeclareRobustCommand{\addCite}[1]{\textcolor{lightgreen}{\mbox{#1}}}

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\cite{higham2002accuracy}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\cite{goubault2006static} hardly scale
to large code bases, particularly in dynamically-typed languages such as Python.

@@

\pytracer detects numerical instabilities by computing summary statistics across
multiple executions perturbed with numerical noise. While \pytracer's
instrumentation, trace aggregation, and visualization work with various types of
numerical noise, we experimented primarily with Monte-Carlo Arithmetic (MCA) as
it is an accurate model for floating-point errors~\cite{verificarlo}.

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis [23], and random
seed analysis [11].  \add{PyTracer is agnostic to the noise model used and therefore
    works for all methods mentioned}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing to simulate
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision. MCA
can be applied in three modes \add{but we only used the} Random Rounding (RR) \add{in our
    experiments} \remove{, Precision Bounding (PB), and full MCA,} which apply
function $inexact$ to the output. \remove{, the inputs, or both, respectively.
    While the RR mode is equivalent to stochastic rounding, the PB mode can also
    identify catastrophic cancellations.} \add{Random Rounding simulates roundoff errors.}

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA.
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

@@

\pytracer stores traces in the pickle format, a binary format to serialize
Python objects. The main advantage of the pickle format compared to other
serialization approaches such as marshal or JSON is the ability to serialize
most Python objects with a native compression. In addition, pickle writes
Python objects sequentially, which preserves the temporal ordering required by
subsequent trace analyses.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. Therefore, it assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from a different
    control flow path and stops the parsing when encountering one.}

\begin{figure*}
    \centering
    \includegraphics[width=0.75\linewidth]{figure/pytracer_layout.pdf}
    \caption{Pytracer visualization layout.}
    \label{fig:visu-layout}
\end{figure*}

@@

\section{Results}

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{Only instrumented the Python interpreter would
    lead to only perturbe pure Python floating-point operations and thus to miss
    operations in compiled codes (i.e. Cython and shared libraries).} We used the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate will be
    the sampling distribution. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample size (3$\simeq$30)
    can give a good estimate, the issue is to capture "rare events" that could make the
    computation strongly diverge from the mean. For now, determining the number of
    MCA samples to use is still an open research question. 
    Recent theoretical works}~\cite{sohier2021confidence}
\add{on the significant digits computation allow computing an associated
    confidence interval depending on the sample size and the underlying distribution of the outputs.}

% \subsection{Results overview}

% Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
% and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
% the tracing and aggregation steps with three possible outcomes: 1) green for
% success, meaning that the execution ended well and no issues were raised by the
% application, 2) red for numerical error in the application, meaning that the
% execution raised a Python error before the end, and 3) orange for errors due to
% the fuzzy noise injection that are not considered numerical errors, meaning that
% the execution raised a Python error but due to a perturbation that should not
% have been introduced (i.e. a floating-point value representing an integer).
% Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
% Rounding mode was entirely successful for 32/40 examples and showed an average
% precision loss of 8 bits. The remaining examples failed due to numerical errors
% in the libraries, which we discuss hereafter. The instrumentation with Full MCA
% is more invasive and reduced the number of successful executions to 12/40. Among
% the 28/40 failed executions, one failed due to an actual numerical error
% (\texttt{Bayesian Ridge Regression} example) and the other ones failed due to
% perturbations that should not have been introduced.

% Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
% the precision for the main outputs of SciPy and scikit-learn examples. We
% measured the precision in significant bits on the final primary example output
% by taking the element-wise mean for non-scalar values and, if the example
% produces many outputs, the average value among all of them. The numerical
% quality across all examples varies from 0 to 52, with an average of 44
% significant bits. For both modes, we note that only two examples failed due to a
% numerical error (\texttt{face\_recognition} for RR and \texttt{Bayesian Ridge
%     Regression} for Full MCA).

@@

        % \tristan{Replace FFT with SciPy FFT, Interpolation with SciPy
        % Interpolation and Optimization with SciPy Optimization, or add a SciPy
        % header to these rows.} \tristan{Nitpick: you could add a color scheme
        % to sigbits (for instance shades of blue), to make the column more
        % easily readable.} \\
        \emph{trace}: outcome of \pytracer tracing, \emph{parse}: outcome of
        \pytracer trace parsing and aggregation, \emph{result}: precision for
        main outputs in significant bits. \valid: successful run, \remove{\warn: error
            from invalid noise injection,} \cross: numerical error or divergent
        traces. \textit{nan} stands for (\texttt{Not-A-Number}). Example names
        are linked to their source code.}

@@

\subsection{Pytracer is applicable to scientific computation libraries of reference}

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with two possible outcomes: 1) green for
success, meaning that the execution ended well and no issues were raised by the
application, and 2) red for numerical error in the application, meaning that the
execution raised a Python error before the end. \remove{, and 3)
    orange for errors due to the fuzzy noise injection that are
    not considered numerical errors, meaning that the execu-
    tion raised a Python error but due to a perturbation that
    should not have been introduced (i.e. a floating-point value
    representing an integer).}
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples and showed an average
precision loss of 8 bits. The remaining examples failed due to numerical errors
in the libraries, which we discuss hereafter. 
\remove{The instrumentation with Full MCA is
    more invasive and reduced the number of successful exe-
    cutions to 12/40. Among the 28/40 failed executions, one
    failed due to an actual numerical error (Bayesian Ridge
    Regression example) and the other ones failed due to
    perturbations that should not have been introduced.}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the precision for the main outputs of SciPy and scikit-learn examples. We
measured the precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produces many outputs, the average value among all of them. The numerical
quality across all examples varies from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

\noindent \remove{\textbf{3.2 Effect of MCA modes}}

\remove{
    Random Rounding perturbs only the result of a floating-point operation while
    Full MCA perturbs both the inputs and the result. Therefore, Full MCA may
    trigger catastrophic cancellations while Random Rounding only introduces
    round-off errors which generally have a lesser impact. Moreover, Random Rounding
    preserves exact operations --- operations where the result can be exactly
    represented at the virtual precision --- while Full MCA does not.}

\remove{\pytracer can only aggregate traces obtained from the same control flow path. In
    particular, program branches triggered by an unstable floating-point comparison
    lead to different control flows and result in parsing errors. For example,
    comparing a residual error or the distance between two points to a strict
    threshold in an iterative scheme generally results in unstable branches. Full
    MCA even perturbs exact values such as integers represented with floating-point
    values, leading to many execution failures. For instance, the \texttt{fft1}
    example from SciPy raised an \texttt{ValueError: operands could not be broadcast
        together with shapes (601,) (600,)} error due to an array shape mismatch. A
    closer inspection showed that one of the arrays involved in the error was
    created with the NumPy function \texttt{linspace} that itself calls function
    \texttt{arange} (Listing~\ref{fig:pyarray_range_code}). Lines 3174-3178 show
    that the function stores the array size in a floating-point value that will be
    perturbed by Full MCA, as shown in Table~\ref{tab:mca_result_linspace},
    resulting in array sizes that fluctuate between N and N+1. It is worth noting
    that 78\% of execution issues encountered in Full MCA mode instrumentation are
    due to this type of ``typing" errors.  
    Those code regions should not be perturbed by MCA since the injected noise does
    not make sense.}

\subsection{\pytracer supports the visualization of compound data type}

\add{PyTracer supports native floating point formats as well as compound data type,
    such as complex floating point values, N-dimensional arrays and arbitrary Python
    classes. For the Python classes, PyTracer recursively parses the class attributes to
    extract scalar and N-dimensional arrays floating-points values.}

Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs} \add{show the
    visualization of the \texttt{fft\_1D} inputs and outputs values. Results are
    complex value encoded on double precision. PyTracer allows visualizing the
    numerical precision of real (blue points) and imaginary (orange points) parts.}
\add{The} top row in Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs}
show the input mean value while the bottom row shows the standard deviation. The
points with low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces a slight noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal noise
introduced by MCA on the inputs is in the order of $10^{-14}$, two orders of
magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 corresponds to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D-array that is natively supported by
    PyTracer. We observe a significant loss of precision in places, which follows
    an interesting pattern that could be further investigated.}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{.3\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figure/spline_2d/bisplev_mean.pdf}
        \caption{Mean value}
        \label{fig:bisplev_mean}
    \end{subfigure}
    \begin{subfigure}{.3\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figure/spline_2d/bisplev_mean_log.pdf}
        \caption{Absolute mean value (log)}
        \label{fig:bisplev_mean_log}
    \end{subfigure}
    \begin{subfigure}{.3\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figure/spline_2d/bisplev_std_log.pdf}
        \caption{Standard deviation (log)}
        \label{fig:bisplev_std_log}
    \end{subfigure}
    \caption{2D Spline interpolation results. Figure~\ref{fig:bisplev_mean}
        shows the mean result of the 2D-spline interpolation.
        Figures~\ref{fig:bisplev_mean_log} and~\ref{fig:bisplev_std_log} show
        respectively in the logarithmic scale the absolute mean value and the
        standard deviation of the 5 samples run within RR mode. 
    }
    \label{fig:spline2d_rr}
\end{figure*}

\subsection{\pytracer allows visualizing numerical stability at a fine grained level}

\add{We illustrate on two examples how PyTracer is effective for analyzing
    the numerical stability at a fine grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the name associated with a face.
\add{When executing within fuzzy environment, this example crashed with the following
    Python error raised: \texttt{"SVD did not converge"}.}
This example is the only one to raise an error due to the
non-convergence of the Singular Value Decomposition (SVD) performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we see that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
makes the SVD to not converge with MCA.

Replacing \texttt{gesdd} for \texttt{gesvd} solves the SVD convergence issue.
\pytracer allows to quickly visualize the new precision of the result.
Figure~\ref{fig:randomized_svd_S} shows the precision of the eigenvalues
computed displayed in decreasing order with \texttt{gesvd} \add{in single precision}, 
which varies between 15 and 20
significant bits out of the 24 available in single precision. The classifier
metrics (precision, recall, F1-score, and support) are identical to the IEEE
execution for all MCA samples. \add{We can note that even with using the \texttt{gesvd},
    the precision of the singular values (vectors) is not uniform. 
    Indeed, we can see that the precision of the lowest singular values (vectors) decreases 
    until reaching 0 significant bits.}
